First Order Alternation

### 2018-09-20 quantifiers/draft.tex
ok

A \emph{sentence} is a formula $\phi$ in which each variable occurs
under the scope of a quantifier, i.e. $\fv{}{\phi} = \emptyset$.  A
formula is \emph{positive} if each predicate symbol occurs under an
even number of negations. A formula is in \emph{prenex form} if it is
of the form $\varphi = Q_1x_1 \ldots Q_nx_n ~.~ \phi$, where $\phi$
has no quantifiers. In this case we call $\phi$ the \emph{matrix} of
$\varphi$. Every first-order formula can be written in prenex form, by
renaming each quantified variable to a unique name and moving the
quantifiers upfront.

@@

A \emph{first-order alternating automaton} (automaton) is a tuple $\A
= \tuple{\Sigma,X,Q,\iota,F,\Delta}$, where: \begin{compactitem}
\item $\Sigma$ is the finite set of input events, 
\item $X$ is a finite set of input variables, 
\item $Q$ is a finite set of predicates denoting control states, 
\item $\iota$ is a sentence defining initial configurations,
\item $F \subseteq Q$ is the set of predicates denoting final states, and 
\item $\Delta$ is a set of \emph{transition rules} of the form
  \(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is
  a predicate, $a \in \Sigma$ is an input event, $X \cap
  \set{y_1,\ldots,y_{\#(q)}} = \emptyset$, $\psi$ is a positive
  formula such that $\fv{}{\phi} \subseteq X \cup
  \set{y_1,\ldots,y_{\#(q)}}$ and each predicate that occurs (under an
  even number of negations) in $\psi$ belongs also to $Q$. 
\end{compactitem}
The quantifiers occurring in the right-hand side formula of a
transition rule are referred to as \emph{transition quantifiers}. The
\emph{size} of $\A$ is defined as $\len{\A} = \len{\iota} +
\sum_{\scriptscriptstyle{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta}}
\len{\psi}$.

@@

Given a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in \Sigma[X]^*$, an
\emph{execution} of $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$ is
a (possibly infinite) forest $\T = \set{T_1,T_2,\ldots}$, where each
$T_i$ is a tree labeled with configurations, such that:
\begin{compactitem}
\item the labels of the roots of $T_1, T_2, \ldots$ form a cube from
  $\cube{\minsem{\iota}}$,
%
\item if $q(d_1,\ldots,d_{\#(q)})$ labels a node on the level $j \in
  [n-1]$ in $T_i$, then the labels of its children form a cube from
  $\cube{\minsem{\psi}_{\nu_{j+1}[y_1 \leftarrow d_1,\ldots,y_{\#(q)}
        \leftarrow d_{\#(q)}]}}$, where \(q(y_1,\ldots,y_{\#(q)})
  \arrow{a_{j+1}(X)}{} \psi \in \Delta\) is a transition rule.
\end{compactitem}
An execution $\T$ over $w$ is said to be \emph{accepting} if and only
if \begin{inparaenum}[(i)]
\item all paths in $\T$ have the same length $n$, and
\item the frontier of each tree $T \in \T$ is labeled with \emph{final
  configurations} $q(d_1,\ldots,d_{\#(q)})$, where $q \in F$.
\end{inparaenum} In this case, we say that $\A$ accepts $w$ and
denote by $\lang{\A}$ the set of words accepted by $\A$.

In this paper, we address the following questions: \begin{compactenum}
\item \emph{boolean closure}: given automata $\A_i =
  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$?
\end{compactenum}

@@

The problem of checking emptiness of a given automata is undecidable,
even for automata with predicates of arity two, whose transition rules
use only equalities and disequalities, with no transition quantifiers
\cite{Farzan15}. Since even such simple classes of alternating
automata have no general decision procedure for emptiness, we use an
abstraction-refinement semi-algorithm based on \emph{lazy annotation}
\cite{McMillan06,McMillan14}. In a nutshell, a lazy annotation
procedure systematically explores the set of execution traces of a
program in search for a counterexample. When a spurious counterexample
is found, the search backtracks and the program path is annotated with
a set of learned facts, that excludes the path from future
searches. These facts are interpolants extracted from the proof of
spuriousness of the given (finite) path. The semi-algorithm uses
moreover a coverage relation to ensure that a spurious counterexample
is never explore twice, thus making progress towards either finding a
real counterexample or a proof of emptiness. In the rest of this
section, we discuss the generation of interpolants
(\S\ref{sec:interpolants}) and the coverage (\S\ref{sec:coverage})
used by our emptiness checking semi-algorithm (\S\ref{sec:impact}).

\subsection{Interpolants}\label{sec:interpolants}

Typically, when checking the unreachability of a set of program
configurations or, more generally, the infeasability of a set of Horn
clauses \cite{BjornerGurfinkelMcMillanRybalchenko15}, the interpolants
used to annotate the unfolded control structure are assertions about
the values of the program variables in a given control state, at a
certain step of a given execution. As we discuss next, it is useful to
distinguish between
\begin{inparaenum}[(i)]
\item locality of interpolants w.r.t. a given control state (control
  locality) and
%
\item locality w.r.t. a given time stamp (time locality).
\end{inparaenum} 
In logical terms, \emph{control-local} interpolants are defined by
formulae involving a single predicate symbol, whereas
\emph{time-local} interpolants involve only predicates $\stamp{q}{i}$
and variables $\stamp{x}{i}$, for a single $i \geq 0$.

@@

\begin{lemma}\label{lemma:subst-quant}
  Given a non-empty input event sequence $\alpha = a_1 \ldots a_n \in
  \Sigma^*$, such that $\accform{\alpha}$ is unsatisfiable, let
  $Q_1x_1 \ldots Q_mx_m ~.~ \widehat{\Phi}$ be a prenex form of
  $\quantform{\alpha}$, where $Q_ix_i$ are transition quantifiers and
  $\xi : [1,m] \rightarrow [n]$ is a monotonic function mapping each
  quantified variable to the time-stamp of the step where it occurs in
  $\accform{\alpha}$. Then the following hold: \begin{compactenum}
  \item\label{it1:subst-quant} there exist witness terms $\tau_{i_1},
    \ldots, \tau_{i_\ell}$, where $1 \leq i_1 < \ldots < i_\ell \leq
    m$ and $\voc{\tau_{i_j}} \subseteq \stamp{X}{\leq \xi(i_j)}$ for
    all $j \in [1,\ell]$, such that $\widehat{\Phi}[\tau_{i_1}/x_{i_1},
      \ldots, \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable, and
    %
  \item\label{it2:subst-quant} there exists a GLI $(I_0, \ldots, I_n)$
    for $\alpha$, such that $\voc{I_k} \cap \stamp{Q}{\leq n}
    \subseteq \stamp{Q}{k}$ and $\voc{I_k} \cap \stamp{X}{\leq n}
    \subseteq \stamp{X}{\leq k}$, for all $k \in [n]$.
  \end{compactenum}
\end{lemma}
\proof{(\ref{it1:subst-quant}) If $\accform{\alpha}$ is unsatisfiable,
  by Lemmas \ref{lemma:quant} and \ref{lemma:subst}, we obtain that,
  successively $\quantform{\alpha}$ and $\substform{\alpha}$ are
  unsatisfiable. Let $Q_1x_1 \ldots Q_mx_m ~.~ \widehat{\Phi}$ and
  $Q_1x_1 \ldots Q_mx_m ~.~ \overline{\Phi}$ be prenex forms for
  $\quantform{\alpha}$ and $\substform{\alpha}$,
  respectively. Further, let $\xi : [1,m] \rightarrow [n]$ be the
  monotone function mapping of each quantified variable to time-stamp
  where it was introduced. Then there exist witness terms $\tau_{i_1},
  \ldots, \tau_{i_\ell}$, where $\set{i_1,\ldots,i_\ell}=\set{i \in
    [1,m] \mid Q_i=\forall}$, such that $\voc{\tau_{i_j}} \subseteq
  \stamp{X}{\xi(i_j)}$, for all $j \in [1,\ell]$ and
  $\overline{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
    \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable.

>>>>

A \emph{sentence} is a formula $\phi$ in which each variable occurs
under the scope of a quantifier, i.e. $\fv{}{\phi} = \emptyset$.  A
formula is \emph{positive} if each predicate symbol occurs under an
even number of negations and we denote by $\posforms(Q,X)$ the set of
positive formulae with predicates from the set $Q \subseteq \preds$
and free variables from the set $X \subseteq \vars$.

A formula is in \emph{prenex form} if it is of the form $\varphi =
Q_1x_1 \ldots Q_nx_n ~.~ \phi$, where $\phi$ has no quantifiers. In
this case we call $\phi$ the \emph{matrix} of $\varphi$. Every
first-order formula can be written in prenex form, by renaming each
quantified variable to a unique name and moving the quantifiers
upfront.

@@

A \emph{first-order alternating automaton} (automaton) is a tuple $\A
= \tuple{\Sigma,X,Q,\iota,F,\Delta}$, where: \begin{compactitem}
\item $\Sigma$ is the finite set of input events, 
\item $X$ is a finite set of input variables, 
\item $Q$ is a finite set of predicates denoting control states, 
\item $\iota \in \posforms(Q,\emptyset)$ is a sentence defining initial configurations,
\item $F \subseteq Q$ is the set of predicates denoting final states, and 
\item $\Delta$ is a set of \emph{transition rules} of the form
  \(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is
  a predicate, $a \in \Sigma$ is an input event and $\psi \in
  \posforms(Q,X \cup \set{y_1,\ldots,y_{\#(q)}})$ is a positive
  formula, where $X \cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$.
\end{compactitem}
The quantifiers occurring in the right-hand side formula of a
transition rule are referred to as \emph{transition quantifiers}. The
\emph{size} of $\A$ is defined as $\len{\A} = \len{\iota} +
\sum_{\scriptscriptstyle{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta}}
\len{\psi}$.

@@

\begin{definition}\label{def:execution}
Given a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in \Sigma[X]^*$ and a cube $c$, an
\emph{execution} of $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$, starting with $c$, is
a (possibly infinite) forest $\T = \set{T_1,T_2,\ldots}$, where each
$T_i$ is a tree labeled with configurations, such that:
\begin{compactenum}
\item\label{it1:execution} $c = \set{T(\epsilon) \mid T \in \T}$ is the set of
  configurations labeling the roots of $T_1,T_2,\ldots$ and
%
\item\label{it2:execution} if $q(d_1,\ldots,d_{\#(q)})$ labels a node on the level $j \in
  [n-1]$ in $T_i$, then the labels of its children form a cube from
  $\cube{\minsem{\psi}_{\eta}}$, where $\eta = \nu_{j+1}[y_1
    \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow d_{\#(q)}]$ and
  \(q(y_1,\ldots,y_{\#(q)}) \arrow{a_{j+1}(X)}{} \psi \in \Delta\) is
  a transition rule of $\A$.
\end{compactenum}
\end{definition}

\begin{definition}\label{def:accepting}
An execution $\T$ over $w$, starting with $c$, is \emph{accepting} if
and only if \begin{compactitem}
\item\label{it1:accepting} all paths in $\T$ have the same length $n$,
  and
  %
\item\label{it2:accepting} the frontier of each tree $T \in \T$ is
  labeled with \emph{final configurations} $q(d_1,\ldots,d_{\#(q)})$,
  where $q \in F$.
\end{compactitem} 
If $\A$ has an accepting execution over $w$ starting with a cube $c
\in \cube{\minsem{\iota}}$, then $\A$ \emph{accepts} $w$ and let
$\lang{\A}$ be the set of words accepted by $\A$.
\end{definition}
In this paper, we address the following questions: \begin{compactenum}
\item \emph{boolean closure}: given automata $\A_i =
  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$?
\end{compactenum}

@@

The problem of checking emptiness of a given automata is undecidable,
even for automata with predicates of arity two, whose transition rules
use only equalities and disequalities, with no transition quantifiers
\cite{Farzan15}. Since even such simple classes of alternating
automata have no general decision procedure for emptiness, we use an
abstraction-refinement semi-algorithm based on \emph{lazy annotation}
\cite{McMillan06,McMillan14}. In a nutshell, a lazy annotation
procedure systematically explores the set of execution traces of a
program in search for a counterexample. When a spurious counterexample
is found, the search backtracks and the program path is annotated with
a set of learned facts, that excludes the path from future
searches. These facts are interpolants extracted from the proof of
spuriousness of the given (finite) path. The emptiness checking
semi-algorithm uses moreover a coverage relation to ensure that a
spurious counterexample is never explore twice, thus making progress
towards either finding a real counterexample or a proof of emptiness.

Our implementation is a version of the \impact~ lazy annotation
semi-algorithm \cite{McMillan06}. The detailed pseudocode has been
given for a simpler model of boolean alternating automata extended
with variables, in \cite{IosifXu18} and is reported, for
self-completeness, in Appendix \ref{app:pseudocode}. For the time
being, we shall content ourselves of a high-level definition of the
data structure used to prove emptiness of an automaton.

Given a finite input event alphabet $\Sigma$, for two sequences
$\alpha, \beta \in \Sigma^*$, we say that $\alpha$ is a prefix of
$\beta$, written $\alpha \prefix \beta$, if $\alpha=\beta\gamma$ for
some sequence $\gamma\in\Sigma^*$. A set $S$ of sequences
is: \begin{compactitem}
\item \emph{prefix-closed} if for each $\alpha \in S$, if $\beta \prefix
\alpha$ then $\beta \in S$, and
\item \emph{complete} if for each $\alpha \in S$, there exists $a \in
  \Sigma$ such that $\alpha a \in S$ if and only if $\alpha b \in S$
  for all $b \in \Sigma$. 
\end{compactitem}
Observe that a prefix-closed set is the backbone of a tree whose edges
are labeled with input events. If the set is, moreover, complete, then
every node of the tree has either zero successors, in which case it is
called a \emph{leaf}, or it has a successor edge labeled with $a$ for
each input event $a \in \Sigma$.

\begin{definition}\label{def:unfolding}
  An \emph{unfolding} of an automaton $\A =
  \tuple{\Sigma,X,Q,\iota,F,\Delta}$ is a finite partial mapping $U :
  \Sigma^* \rightharpoonup_{\mathit{fin}} \posforms(Q,X)$, such
  that: \begin{compactenum}
  \item\label{it1:unfolding} $\dom(U)$ is a finite prefix-closed
    complete set,
    %
  \item\label{it2:unfolding} $U(\epsilon) = \iota$, and
    %
  \item\label{it3:unfolding} for each sequence $\alpha a \in \dom(U)$,
    such that $\alpha \in \Sigma^*$ and $a \in \Sigma$:
    \[\stamp{U(\alpha)}{0} \wedge
    \bigwedge_{q(\vec{y}) \arrow{a(X)}{} \psi} \forall y_1 \ldots
    \forall y_{\#q} ~.~ \stamp{q}{0}(\vec{y}) \arrow{a(X)}{}
    \stamp{\psi}{1} \models \stamp{U(\alpha a)}{1}\]
  \end{compactenum}
  Moreover, $U$ is \emph{safe} if for each $\alpha \in \dom(U)$, the
  formula $U(\alpha) \wedge \bigwedge_{q \in Q \setminus F} \forall
  y_1 \ldots \forall_{y_{\#(q)}} ~.~ q(\vec{y}) \rightarrow \bot$ is
  unsatisfiable. 
\end{definition}

Lazy annotation semi-algorithms \cite{McMillan06,McMillan14}
essentially build unfoldings of automata trying to discover
counterexamples for emptiness. If the automaton $\A$ in question is
non-empty, a systematic enumeration of the input event
sequences\footnote{For instance, using breadth-first search.} from
$\Sigma^*$ will suffice to discover a word $w \in \lang{\A}$, provided
that the first-order theory of the data domain $\Data$ is decidable
(Lemma \ref{lemma:acceptance}). However, if $\lang{\A} = \emptyset$,
the enumeration of input event sequences may, in principle, run
forever. The typical way of fighting this divergence problem is to
define a \emph{coverage} relation between the nodes of the unfolding
tree.

\begin{definition}\label{def:coverage}
  Given an unfolding $U$ of an automaton $\A =
  \tuple{\Sigma,X,Q,\iota,F,\Delta}$ a node $\alpha \in \dom(U)$ is
  \emph{covered} by another node $\beta \in \dom(U)$, denoted $\alpha
  \cover \beta$, if and only if there exists a node $\alpha' \prefix
  \alpha$ such that, for all words $w \in \Sigma[X]^*$ and all cubes
  $c \in \cube{\sem{U(\alpha')}}$, if $\A$ accepts $w$ started in $c$
  then there exists a cube $d \in \cube{\sem{U(\beta)}}$ such that
  $\A$ accepts $w$ started in $d$. Moreover, $U$ is \emph{closed} if
  and only if every leaf from $\dom(U)$ is covered.
\end{definition}

A lazy annotation semi-algorithm can stop and report emptiness if it
succeeds in building a closed and safe unfolding of the automaton. For
instance, the \impact~ semi-algorithm \cite{McMillan06}

The higher complexity of the first-order alternating automata model
versus the automata considered in \cite{IosifXu18}, which uses boolean
constants for control states and whose transition rules have no
quantifiers, manifests itself within the method used to generate the
labels of the unfolding, which is discussed in the rest of this
section.

\subsection{Interpolant Generation}\label{sec:interpolants}

Typically, when checking the unreachability of a set of program
configurations or, more generally, the infeasability of a set of Horn
clauses \cite{BjornerGurfinkelMcMillanRybalchenko15}, the interpolants
used to annotate the unfolded control structure are assertions about
the values of the program variables in a given control state, at a
certain step of a given execution. It is useful to distinguish between
\begin{inparaenum}[(i)]
\item locality of interpolants w.r.t. a given control state (control
  locality) and
%
\item locality w.r.t. a given time stamp (time locality).
\end{inparaenum} 
In logical terms, \emph{control-local} interpolants are defined by
formulae involving a single predicate symbol, whereas
\emph{time-local} interpolants involve only predicates $\stamp{q}{i}$
and variables $\stamp{x}{i}$, for a single $i \geq 0$.

@@

\begin{lemma}\label{lemma:subst-quant}
  Given a non-empty input event sequence $\alpha = a_1 \ldots a_n \in
  \Sigma^*$, such that $\accform{\alpha}$ is unsatisfiable, let
  $Q_1x_1 \ldots Q_mx_m ~.~ \widehat{\Phi}$ be a prenex form of
  $\quantform{\alpha}$, where $Q_ix_i$ are transition quantifiers and
  $\xi : [1,m] \rightarrow [n]$ is a monotonic function mapping each
  quantified variable to the time-stamp of the step where it occurs in
  $\accform{\alpha}$. Then the following hold: \begin{compactenum}
  \item\label{it1:subst-quant} there exist witness terms $\tau_{i_1},
    \ldots, \tau_{i_\ell}$, where $1 \leq i_1 < \ldots < i_\ell \leq
    m$ and $\voc{\tau_{i_j}} \subseteq \stamp{X}{\leq \xi(i_j)}$ for
    all $j \in [1,\ell]$, such that $\widehat{\Phi}[\tau_{i_1}/x_{i_1},
      \ldots, \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable, and
    %
  \item\label{it2:subst-quant} one can effectively build a GLI $(I_0,
    \ldots, I_n)$ for $\alpha$, such that $\voc{I_k} \cap
    \stamp{Q}{\leq n} \subseteq \stamp{Q}{k}$ and $\voc{I_k} \cap
    \stamp{X}{\leq n} \subseteq \stamp{X}{\leq k}$, for all $k \in
          [n]$.
  \end{compactenum}
\end{lemma}
\proof{(\ref{it1:subst-quant}) If $\accform{\alpha}$ is unsatisfiable,
  by Lemmas \ref{lemma:quant} and \ref{lemma:subst}, we obtain that,
  successively $\quantform{\alpha}$ and $\substform{\alpha}$ are
  unsatisfiable. Let $Q_1x_1 \ldots Q_mx_m ~.~ \widehat{\Phi}$ and
  $Q_1x_1 \ldots Q_mx_m ~.~ \overline{\Phi}$ be prenex forms for
  $\quantform{\alpha}$ and $\substform{\alpha}$,
  respectively. Further, let $\xi : [1,m] \rightarrow [n]$ be the
  monotone function mapping of each quantified variable to time-stamp
  where it was introduced. Then there exist witness terms $\tau_{i_1},
  \ldots, \tau_{i_\ell}$, where $\set{i_1,\ldots,i_\ell}=\set{i \in
    [1,m] \mid Q_i=\forall}$, such that $\voc{\tau_{i_j}} \subseteq
  \stamp{X}{\xi(i_j)}$, for all $j \in [1,\ell]$ and
  $\overline{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
    \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable.
